Analyse and data mining of a chromatograph during drilling operations

### 2018-12-02 papers/chromatography (for ethics course)/paper.tex
Finished chroma paper for ethics course.

    Gas chromatography means to separate components such that they can be analyzed independently. To achive this a
    separation column is used. A probe of gas, taken from drilling installation, is pushed through the column which
    means that all components enter into it at the same time. The column is filled with carefully choosen particles,
    whose type and size are competitive commercial advantages (and because this is not the object of the current paper,
    their details will not be disclosed). The probe is pushed through the column using a bearing gas, usually hydrogen.
    The probe has to make its way through the column's particles and because of the friction between them different
    components will have different speeds with which they advance. Thus, the components are reaching the end of the 
    column at different times. The time which is needed for a component to traverse the column is called retention time.
    Figure 1 contains a representation of a column and how two components, A and B, enter into the column at the same
    time but they exit at different times.

@@

    \begin{figure}
        \centering
        \includegraphics[width=0.5\textwidth]{column.png}
        \caption{Gas separation column}
    \end{figure}

@@

    The components are slowly (or fast, depending on the performance of the chromatograph) pouring out of the column
    into a detector called Flame Ionisation Detector or FID. Its role is to transform over the analyses time the quantity
    of each component into voltage. This voltage is read by ADC boards and transmitted to a computer for further calculations.
    Figure 2 shows the schematics of such a FID. Item 1 represents the separation column, item 2 represents suplimentary
    gas supply, 3 represents hydrogen supply, 4 represents the lighter, 5 represents the collector electrode, 6 is the
    polarizing electrode, 7 is the burner and 8 is the air supply.

    \begin{figure}
        \centering
        \includegraphics[width=0.5\textwidth]{fid.png}
        \caption{Flame Ionisation Detector}
    \end{figure}

@@

@@

    \begin{figure}
        \centering
        \includegraphics[width=0.5\textwidth]{fidgraph.png}
        \caption{FID Graph}
    \end{figure}

>>>>

    Gas chromatography means to separate components such that they can be analyzed independently. To achive this a
    separation column is used. A probe of gas, taken from drilling installation, is pushed through the column which
    means that all components enter into it at the same time. The column is filled with carefully choosen particles,
    whose type and size are competitive commercial advantages (and because this is not the object of the current paper,
    their details will not be disclosed). The probe is pushed through the column using a bearing gas, usually hydrogen.
    The probe has to make its way through the column's particles and because of the friction between them different
    components will have different speeds with which they advance. Thus, the components are reaching the end of the 
    column at different times. The time which is needed for a component to traverse the column is called retention time.
    Figure \ref{fig:column} contains a representation of a column and how two components, A and B, enter into the column at the same
    time but they exit at different times.

@@

    \begin{figure}
        \centering
        \includegraphics[width=0.5\textwidth]{column.png}
        \caption{Gas separation column}
        \label{fig:column}
    \end{figure}

@@

    The components are slowly (or fast, depending on the performance of the chromatograph) pouring out of the column
    into a detector called Flame Ionisation Detector or FID. Its role is to transform over the analyses time the quantity
    of each component into voltage. This voltage is read by ADC boards and transmitted to a computer for further calculations.
    Figure \ref{fig:fid} shows the schematics of such a FID. Item 1 represents the separation column, item 2 represents suplimentary
    gas supply, 3 represents hydrogen supply, 4 represents the lighter, 5 represents the collector electrode, 6 is the
    polarizing electrode, 7 is the burner and 8 is the air supply.

    \begin{figure}
        \centering
        \includegraphics[width=0.5\textwidth]{fid.png}
        \caption{Flame Ionisation Detector}
        \label{fig:fid}
    \end{figure}

@@

    \begin{figure}
        \centering
        \includegraphics[width=0.5\textwidth]{fidgraph.png}
        \caption{FID Graph}
        \label{fig:fidgraph}
    \end{figure}

@@

    The mathematical model is so simple, one could wonder if it works at all: it's based on the least squares equation.
    From figure \ref{fig:fidgraph} it can be seen that during one minute of analyses the first two FIDs have some components present.
    Using different rules for establishing a baseline the area of each componente (each "hill") is computed. This area is 
    proportional with the amount of the component from the last probe analysed. The engineer who builds the chromatograph and then
    the engineer who maintain it will introduce, in consecutive analyses, different gases with a precisely known composition, using
    different concentrations. Thus, a series of analyses will be made, knowing for each one the calibrating gas composition and the
    results. A look tabel is born from this operation.

    During normal operations the look up tabel will be used to analyze gases from drilling site. To be able to give results for any
    given concentration, for any given component, an interpolation is needed between two known points, using first two points or
    using last two points. This kind of interpolation is very similar to an average between two numbers. Equation \ref{eq:average}
    represents such an average: \(K_n\) and \(K_{n+1}\) are numerical integrations results and  \(C_n\) and \(C_{n+1}\) are
    corresponding component's concentrations from the lookup table. \(R\) will be the result of the current analyze.

    \begin{equation}
        R = \frac{C_n + C_{n+1}}{K_n + K_{n+1}}, K_{n} < R < K_{n+1}
        \label{eq:average}
    \end{equation}

    Another method for obtaining results using the data from the lookup table is to use the least squares method. Equation \ref{eq:leastSquares}
    shows the gist of this method. \(x_{1}\) through \(x_{n}\) are the known concentrations from calibration operation. \(y\) terms are
    the numerical integration results for a component, for each concentration. \(b\) terms are the coefficients which are found by 
    solving the equation \ref{eq:leastSquares}. \(b\) will form a polynom which will be used during normal analyze mode to computed
    the real concentration of a component from a probe taken from the drilling site.

    \begin{equation}
        \begin{bmatrix}
             x_0 \\ \vdots \\ x_{n} 
         \end{bmatrix}
             =
         \begin{pmatrix}
            y_{00} & \hdots & y_{0(m-1)} & y_{0m}\\
            \vdots & & \vdots & \vdots \\
            y_{n0} & \hdots & y_{n(m-1)} & y_{nm}
            \end{pmatrix}
            \times
            \begin{pmatrix}
            b_0 \\ \vdots \\ b_m
        \end{pmatrix}
        \label{eq:leastSquares}
    \end{equation}
